Modular Model-Checking of a Byzantine Fault-Tolerant Protocol

### 2016-12-08 doc/om-paper/paper.tex
pretty bar for agreement, validity.
more edits.

\newboolean{submission}  %set to true for the submission version
%\setboolean{submission}{false}
\setboolean{submission}{true}
\ifthenelse
{\boolean{submission}}
{ %
  \newcommand{\lee}[1]{ } %
  \newcommand{\ben}[1]{ } %
  % Put your own TODO macros here and below
} %hide todo
{ %
  \newcommand{\lee}[1]{ {\color{blue}$<$lee: #1$>$} } %
  \newcommand{\ben}[1]{ {\color{purple}$<$ben: #1$>$} } %
  \usepackage{hyperref} %
  \usepackage[inline]{showlabels} %
}

@@

\begin{abstract}
With proof techniques like IC3 and $k$-induction, model-checking scales further
than ever before.  Still, fault-tolerant distributed systems are particularly
challenging to model-check given their large state spaces and
non-determinism. The typical approach to controlling complexity is to construct
ad-hoc abstractions of faults, message-passing, and behaviors.  However, these
abstractions come at the price of divorcing the model from its implementation
and making refactoring difficult. In this work, we present a framework for
fault-tolerant distributed system verification that combining ideas from the literature including calendar automata,
symbolic fault injection, and abstract state machines observers, and then we apply the framework to model-checking various implementations of the Hybrid Oral Messages algorithm that differ in the fault model, timing model, and local node behavior. We show that despite being implementation-level models, the verifications are scalable and modular, insofar as isolated changes to an implementation require modular changes to the model and proofs.
This work is carried out in SAL model-checker.
\end{abstract}

\lee{grep for synchronous observers section references}
\lee{grep for ASM vs abstract transition system}

@@

In this paper, we present a framework for modeling fault-tolerant distributed systems, and use that framework to verify several variant implementations of the Byzantine fault-tolerant Hybrid Oral Messages algorithm (\OMH)~\cite{Lincoln-Rushby}. The framework combines ideas from the literature and folklore to build scalable and modular formal models suitable for infinite-state model-checking, and it reduces the need for ad-hoc abstractions and optimizations while maintaining scalability. In Section~\ref{sec:model}, we present the important aspects of the framework, including calendar automata, originally developed by Dutertre and Sorea~\cite{cal}, symbolic fault-injection, and abstract state machines for verification.

@@

We use the framework to verify implementation-level models of \OMH in which message passing is explicit, nodes are not enforced to execute strictly synchronously, and voting is explicit. In short, the models corresponds closely with an implementation of the algorithm. In Section~\ref{sec:byz}, we first describe \OMH, then an implementation of it that uses the Boyer-Moore Fast Majority Vote algorithm~\cite{mjrty}. We then describe a set of modular invariants, such that the invariants only concern specific aspects of the model (e.g., faults, local node behavior, or the passage of time). The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm.

@@

@@

% ------------------------------------------------------------
\section{Formal Framework}\label{sec:model}
Here we describe our formal framework specialized for fault-tolerant distributed systems. The framework draws on three principal abstractions: calendar automata, symbolic fault injection, and abstract state machines observers; we describe each below.

@@

The event of receiving a message initiates a process to update its local state machine and generate additional messages to send. When the process is updating its local state machine, the event calendar is paused. That is, updating an event $recv(i, j, t, m)$ also includes updating $j$'s state machine.

@@

\lee{Cutting synchronous observer section; need to talk about use for lemmas elsewhere.}

@@

$\OMH$ is designed to ensure \emph{validity} and \emph{agreement} properties under suitable hypotheses on the number and type of faults in the system. Validity states that if the general is nonfaulty, then every lieutenant outputs the value sent by the general. Agreement states that each lieutenant outputs the same value. More formally, Let $l_i, l_j$ denote the outputs of lieutenants $i, j \in L$, respectively, and let $v$ be the value the general broadcasts:
\noindent
\begin{multicols}{2}
\begin{equation}
  \tag{Validity}
    \forall \,i. \quad v \neq \ERR \Rightarrow l_i = v
\end{equation}\break
\begin{equation}
  \tag{Agreement}
    \forall \,i, j. \quad l_i = l_j
\end{equation}
\end{multicols}
%

@@

\lee{we need to mention MJRTY here or in the modular verifiation section}

@@

@@

The \texttt{observer} is a synchronous observer~\cite{Rushby:SAS14} that encodes the validity and agreement properties as state machines. State variables denoting validity and agreement are set to be false if the receivers have completed their vote but the respective properties do not hold.

Finally, the \texttt{abstractor} and \texttt{abstract\_monitor} encode an abstract state machine observer for the system, as described in Section~\ref{sec:abstract}.

@@

    \node[state] (calendar) {calendar (1)};
    \node[state] (lieutenant) [right=1.75cm of calendar] {\begin{tabular}{c}lieutenant\\state-machine (2)\end{tabular}};
    \node[state] (faults) [below=0.5cm of calendar] {faults (5)};
    \node[state] (asm) [right=0.5cm of faults] {ASM (1)};
    \node[state] (voting) [below=0.5cm of lieutenant] {voting (2)};

@@

To make the proof scalable, we specify inductive invariants to be used by SAL's $k$-induction engine. There are 11 invariants, falling into five categories:
\begin{enumerate}
  \item \emph{Calendar automata}: Lemmas relating to the calendar automata model. These include lemmas such as time being monotonic, channels missing messages if there is no calendar event, and only nodes associated with a calendar event may execute their local state machines.
  \item \emph{Abstract state machine (ASM)}: Lemmas relating the ASM synchronous observer states to the implementation states.
  \item \emph{Receiver local behavior}: Lemmas describing the modes of behavior of the receivers. The major modes of the state machine are receiving messages, then once it has filled its buffer, it votes, and after voting returns the result. One additional lemma notes that the messages currently received plus missing messages equals the total number of expected messages.
  \item \emph{Faults}: Lemmas characterizing the effect of a fault in a single broadcast. Examples include lemmas stating that if a node receives a faulty message, some ``upstream'' node in the communication path was faulty. Another example is that the faults of messages latched by a node in its buffer match the faults ascribed to the sender in the calendar event.
  \item \emph{Voting}: Lemmas proving that the Fast Majority Vote algorithm implements a majority vote, if one exists. These lemmas are nearly verbatim transcriptions from the journal proofs~\cite{mjrty}.
\end{enumerate}
\noindent
The proof structure is shown in Figure~\ref{fig:proof}. The number of lemmas per category are shown in parentheses. Arrows denote dependencies. For example, the ASM lemmas depend on both the calendar automata and receiver state-machine lemmas. As can be seen, the proof structure is modular. The calendar lemmas are general and independent of any particular protocol or fault model. Similarly, lemmas about the internal behavior of a receiver is independent of the global protocol behavior. It is also independent of the effect of faults on the system---the only ``knowledge'' of faults that receiver has is whether a fault is benign or not. Lemmas about the behavior of faults in the system are also independent of the particular protocol being modeled. Likewise, lemmas about the particular voting algorithm used depend only on the receiver's internal behavior. Only the ASM depends on both calendar-specific and local state-machine results, since it is an abstraction of the entire system implementation. Recall, however, that the ASM is a convenience for debugging and can be elided.

@@

\begin{figure}
  \centering
  \begin{tabular}{|p{3cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
  \hline
                                  & Transition systems\linebreak (7 total)  &Definitions\linebreak (58 total) & Invariants\linebreak (11 total) & Invariant classes\linebreak (5 total) \\
  \hline \hline
  %% No Fault Model                  & receivers & 8 modified & 2 modified & ASM, faults \\
  %% \hline
  Omissive Asymmetric Faults       & none      & 1 new, 2 modified & 2 modified & faults             \\
  \hline
  Time-Triggered Messaging        & source, relays, receivers, ASM & 3 new & 2 modified, 3 modified & calendar, faults \\
  \hline
  Mid-Value Selection & receivers & 4 new, 3 modified & 2 modified & ASM, voting \\
  \hline
  \end{tabular}
  \caption{Refactoring effort for protocol modifications, measured by which portions of the model have to be modified.}
  \label{fig:effort}
\end{figure}

@@

Changing the model to be time-triggered principally requires making the source, relays, and receivers driven explicitly by the passage of time. As well, a ``receive window'' is defined at which messages from non-faulty nodes should be received. Messages received outside the window are marked as coming from manifest-faulty senders. To implement a time-triggered model, we create three definitions; one encodes nondeterministic message delay, and two are small helpers called by invariants. The guards in the relays and receivers are modified to latch messages received outside the specified windows as being manifest faults. Additionally, the ASM definition is modified to track the times in the calendar, not just the messages. Two new calendar invariants are introduced, stating that the calendar messages are either empty, or their time-stamps fall within the respective message windows. Then, three invariants classifying faults are relaxed to allow for the possibility of faulty nodes sending benign messages.

\paragraph{Mid-Value Selection.}
Our $\OMH(1)$ model leverages a majority vote in order to tolerate faults. Another choice for the fault masking algorithm used is mid-value selection. This choice is common in applications involving hardware, signal selection, or cases where information about congruence is useful. To implement mid-value selection in our model, we allow messages sent to take values in $\mathbb{R}$ and the receiver transition system is modified in two ways. First, a second buffer is introduced which will hold the sorted contents of the main buffer once voting has commenced. Second, a mid-value select function is called on the sorted buffer and the result is stored as the receiver's vote. The only invariants needing modification were the ASM definition (to account for the values stored by the new buffer and the relation between it and the main buffer) and the voting invariant.

@@

The lemmas described in Section~\ref{sec:invariants} are constructed by-hand and represent multiple days of effort, but that effort includes both model and protocol construction and generalization as well as verification. The counterexamples returned by SAL are very useful for strengthening invariants, but tedious to analyze---a model with five relays and two receivers contains 90 state variables, and there are known counterexamples to models that size~\cite{Lincoln-Rushby}. Once we developed the ASM synchronously-composed observer, the verification effort was sped up considerably.

@@

The Oral Messages algorithm and its variants and its variants have a long history of formal verification. \OM{1} was verified in both the PVS and ACL2 interactive theorem-provers~\cite{Young97:IC}. Also in ACL2, an implementation of a circuit design to implement \OM{1} is given~\cite{om-acl2-impl}; the low-level model most closely relates to our level of detail. A refinement-based verification approach is used, and \OM{1} is specialized to a four-node system. It appears that a general purpose voting algorithm (like Fast MJRTY) is not verified as part of the work. It is not clear how difficult it would be to model and verify changes to the implementation in that framework.

@@

@@

Beyond building an ADSL, another avenue of research is producing a formal proof that a software implementation satisfies the node specification in our framework. While our model of node behavior is low-level, there are gaps. For example, our work is in SAL's language of guarded commands~\cite{SAL} and needs to be either refined or verified to be equivalent to a software implementation's semantics. Another aspect is that behavior related to networking and benign fault diagnosis (e.g., computing a CRC) is are left abstract. In a software implementation, the $send$ and $recv$ functions would have to implement these behaviors.

@@

\lee{link to all results in github}
\lee{talk about compositional verification of properties, even though SAL doesn't directly support it---maybe not so important}
\lee{talk about k-induction vs. PDR---actually I don't think we have much to say}
\lee{talk about lack of axiomatization in model-checking (multiple rushby bugs), but  tradeoff of deadlock. see proglema paper}

>>>>

\newboolean{submission}  %set to true for the submission version
\setboolean{submission}{false}
%\setboolean{submission}{true}
\ifthenelse
{\boolean{submission}}
{ %
  \newcommand{\lee}[1]{ } %
  \newcommand{\ben}[1]{ } %
  % Put your own TODO macros here and below
} %hide todo
{ %
  \newcommand{\lee}[1]{ {\color{blue}$<$lee: #1$>$} } %
  \newcommand{\ben}[1]{ {\color{purple}$<$ben: #1$>$} } %
  \usepackage{hyperref} %
  \usepackage[inline]{showlabels} %
}

@@

\begin{abstract}
With proof techniques like IC3 and $k$-induction, model-checking scales further
than ever before.  Still, fault-tolerant distributed systems are particularly
challenging to model-check given their large state spaces and
non-determinism. The typical approach to controlling complexity is to construct
ad-hoc abstractions of faults, message-passing, and behaviors.  However, these
abstractions come at the price of divorcing the model from its implementation
and making refactoring difficult. In this work, we present a framework for
fault-tolerant distributed system verification that combining ideas from the literature including calendar automata,
symbolic fault injection, and abstract transition systems, and then we apply the framework to model-checking various implementations of the Hybrid Oral Messages algorithm that differ in the fault model, timing model, and local node behavior. We show that despite being implementation-level models, the verifications are scalable and modular, insofar as isolated changes to an implementation require modular changes to the model and proofs.
This work is carried out in SAL model-checker.
\end{abstract}

@@

In this paper, we present a framework for modeling fault-tolerant distributed systems, and use that framework to verify several variant implementations of the Byzantine fault-tolerant Hybrid Oral Messages algorithm (\OMH)~\cite{Lincoln-Rushby}. The framework combines ideas from the literature and folklore to build scalable and modular formal models suitable for infinite-state model-checking, and it reduces the need for ad-hoc abstractions and optimizations while maintaining scalability. In Section~\ref{sec:model}, we present the important aspects of the framework, including calendar automata, originally developed by Dutertre and Sorea~\cite{cal}, symbolic fault-injection, and abstract transition systems for verification.

@@

We use the framework to verify implementation-level models of \OMH in which message passing is explicit, nodes are not enforced to execute strictly synchronously, and voting is explicit. In short, the models corresponds closely with an implementation of the algorithm. In Section~\ref{sec:byz}, we first describe \OMH, then an implementation of it that uses the Boyer-Moore Fast Majority Vote algorithm (Fast~MJRT)~\cite{mjrty}. We then describe a set of modular invariants, such that the invariants only concern specific aspects of the model (e.g., faults, local node behavior, or the passage of time). The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm.

@@

The models and experiments reported herein can be found at \url{https://github.com/GaloisInc/mmc-paper}.

@@

% ------------------------------------------------------------
\section{Formal Framework}\label{sec:model}
Here we describe our formal framework specialized for fault-tolerant distributed systems. The framework draws on three principal abstractions: calendar automata, symbolic fault injection, and abstract transition sytems; we describe each below.

@@

The event of receiving a message initiates a process to update its local transition system and generate additional messages to send. When the process is updating its local transition system, the event calendar is paused. That is, updating an event $recv(i, j, t, m)$ also includes updating $j$'s transition system.

@@

@@

$\OMH$ is designed to ensure \emph{validity} and \emph{agreement} properties under suitable hypotheses on the number and type of faults in the system. Validity states that if the general is nonfaulty, then every lieutenant outputs the value sent by the general. Agreement states that each lieutenant outputs the same value. More formally, Let $l_i, l_j$ denote the outputs of lieutenants $i, j \in L$, respectively, and let $v$ be the value the general broadcasts:
\noindent
\begin{center}
\begin{tabular}{l|l}
\begin{minipage}[t]{0.5\linewidth}
\begin{equation*}
%  \tag{Validity}
    \forall \,i. \quad l_i = v \textnormal{ (Validity)}
\end{equation*}
\end{minipage}&
\begin{minipage}[t]{0.5\linewidth}
\begin{equation*}
%  \tag{Agreement}
    \forall \,i, j. \quad l_i = l_j \textnormal{ (Agreement)}
\end{equation*}
\end{minipage}
\end{tabular}
\end{center}
%

@@

@@

The relays and receivers explicitly send and receive messages and store them in local buffers as needed. In addition, the receivers implement the Fast~MJRY algorithm~\cite{mjrty}.

@@

The \texttt{observer} is a synchronous observer~\cite{Rushby:SAS14} that encodes the validity and agreement properties as synchronously-composed transition systems. State variables denoting validity and agreement are set to be false if the receivers have completed their vote but the respective properties do not hold.

Finally, the \texttt{abstractor} and \texttt{abstract\_monitor} encode an abstract transition system for the system, as described in Section~\ref{sec:abstract}.

@@

    \node[state] (calendar) {calendar (1)};
    \node[state] (lieutenant) [right=1.75cm of calendar] {\begin{tabular}{c}receiver\\local behavior (2)\end{tabular}};
    \node[state] (faults) [below=0.5cm of calendar] {faults (5)};
    \node[state] (asm) [right=0.5cm of faults] {ATS (1)};
    \node[state] (voting) [below=0.5cm of lieutenant] {voting (2)};

@@

To make the proof scalable, we specify inductive invariants to be used by SAL's $k$-induction engine. There are 11 invariants, falling into five categories:
\begin{enumerate}
  \item \emph{Calendar automata}: Lemmas relating to the calendar automata model. These include lemmas such as time being monotonic, channels missing messages if there is no calendar event, and only nodes associated with a calendar event may execute their local transition systems.
  \item \emph{Abstract transition sytem (ATS)}: Lemmas relating the ATS states to the implementation states.
  \item \emph{Receiver local behavior}: Lemmas describing the modes of behavior of the receivers. The major modes of their behavior are receiving messages, then once it has filled its buffer, it votes, and after voting returns the result. An additional lemma notes that the messages currently received plus missing messages equals the total number of expected messages.
  \item \emph{Faults}: Lemmas characterizing the effect of a fault in a single broadcast. Examples include lemmas stating that if a node receives a faulty message, some ``upstream'' node in the communication path was faulty. Another example is that the faults of messages latched by a node in its buffer match the faults ascribed to the sender in the calendar event.
  \item \emph{Voting}: Lemmas proving that the Fast~MJRTY algorithm implements a majority vote, if one exists. These lemmas are nearly verbatim transcriptions from the journal proofs for the algorithm~\cite{mjrty}.
\end{enumerate}
\noindent
The proof structure is shown in Figure~\ref{fig:proof}. The number of lemmas per category are shown in parentheses. Arrows denote dependencies. For example, the ATS lemmas depend on both the calendar automata and receiver state-machine lemmas. As can be seen, the proof structure is modular. The calendar lemmas are general and independent of any particular protocol or fault model. Similarly, lemmas about the internal behavior of a receiver is independent of the global protocol behavior. It is also independent of the effect of faults on the system---the only ``knowledge'' of faults that receiver has is whether a fault is benign or not. Lemmas about the behavior of faults in the system are also independent of the particular protocol being modeled. Likewise, lemmas about the particular voting algorithm used depend only on the receiver's internal behavior. Only the ATS depends on both calendar-specific and local state-machine results, since it is an abstraction of the entire system implementation. Recall, however, that the ATS is a convenience for debugging and can be elided.

@@

\begin{figure}
  \centering
  \begin{tabular}{|p{3cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
  \hline
                                  & Transition systems\linebreak (7 total)  &Definitions\linebreak (58 total) & Invariants\linebreak (11 total) & Invariant classes\linebreak (5 total) \\
  \hline \hline
  %% No Fault Model                  & receivers & 8 modified & 2 modified & ATS, faults \\
  %% \hline
  Omissive Asymmetric Faults       & none      & 1 new, 2 modified & 2 modified & faults             \\
  \hline
  Time-Triggered Messaging        & source, relays, receivers, ATS & 3 new & 2 modified, 3 modified & calendar, faults \\
  \hline
  Mid-Value Selection & receivers & 4 new, 3 modified & 2 modified & ATS, voting \\
  \hline
  \end{tabular}
  \caption{Refactoring effort for protocol modifications, measured by which portions of the model have to be modified.}
  \label{fig:effort}
\end{figure}

@@

Changing the model to be time-triggered principally requires making the source, relays, and receivers driven explicitly by the passage of time. As well, a ``receive window'' is defined at which messages from non-faulty nodes should be received. Messages received outside the window are marked as coming from manifest-faulty senders. To implement a time-triggered model, we create three definitions; one encodes nondeterministic message delay, and two are small helpers called by invariants. The guards in the relays and receivers are modified to latch messages received outside the specified windows as being manifest faults. Additionally, the ATS definition is modified to track the times in the calendar, not just the messages. Two new calendar invariants are introduced, stating that the calendar messages are either empty, or their time-stamps fall within the respective message windows. Then, three invariants classifying faults are relaxed to allow for the possibility of faulty nodes sending benign messages.

\paragraph{Mid-Value Selection.}
Our $\OMH(1)$ model leverages a majority vote in order to tolerate faults. Another choice for the fault masking algorithm used is mid-value selection. This choice is common in applications involving hardware, signal selection, or cases where information about congruence is useful. To implement mid-value selection in our model, we allow messages sent to take values in $\mathbb{R}$ and the receiver transition system is modified in two ways. First, a second buffer is introduced which will hold the sorted contents of the main buffer once voting has commenced. Second, a mid-value select function is called on the sorted buffer and the result is stored as the receiver's vote. The only invariants needing modification were the ATS definition (to account for the values stored by the new buffer and the relation between it and the main buffer) and the voting invariant.

@@

The lemmas described in Section~\ref{sec:invariants} are constructed by-hand and represent multiple days of effort, but that effort includes both model and protocol construction and generalization as well as verification. The counterexamples returned by SAL are very useful for strengthening invariants, but tedious to analyze---a model with five relays and two receivers contains 90 state variables, and there are known counterexamples to models that size~\cite{Lincoln-Rushby}. Once we developed the synchronously-composed ATS observer, the verification effort was sped up considerably.

@@

The Oral Messages algorithm and its variants and its variants have a long history of formal verification. \OM{1} was verified in both the PVS and ACL2 interactive theorem-provers~\cite{Young97:IC}. Also in ACL2, an implementation of a circuit design to implement \OM{1} is given~\cite{om-acl2-impl}; the low-level model most closely relates to our level of detail. A refinement-based verification approach is used, and \OM{1} is specialized to fixed number of nodes.

@@

Our work was heavily influenced by previous verifications of fault-tolerant systems in SAL~\cite{Rushby:OM1,cal,timed-systems}.

@@

Beyond building an ADSL, another avenue of research is producing a formal proof that a software implementation satisfies the node specification in our framework. While our model of node behavior is low-level, there are gaps. For example, our work is in SAL's language of guarded commands~\cite{SAL} and needs to be either refined or verified to be equivalent to a software implementation's semantics. Another aspect is that behavior related to networking and benign fault diagnosis (e.g., computing a CRC) is left abstract. In a software implementation, the $send$ and $recv$ functions would have to implement these behaviors.

@@
